---
title: "Assignment 1"
author: "Group 69"
date: "24.02.2023"
output: 
  pdf_document: 
    latex_engine: lualatex
fontsize: 11pt
highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

The data set birthweight.txt contains the birthweights (in grams) of 188
newborn babies. Denote the underlying mean birthweight by mu.

a)  Check normality of the data. Assuming normality (irrespective of
    your conclusion about normality), construct a bounded 96%-CI for mu.
    Evaluate the sample size needed to provide that the length of the
    96%-CI is at most 100. Compute a bootstrap 96%-CI for mu and compare
    it to the above CI.

To start with we import the birthweight.txt dataset and store it to the
respective variable.

To make an assumption about the distribution we display the data using
the following function.

```{r}
dataset = read.table("birthweight.txt", header = T)
hist(dataset$birthweight, main="Histogram of Birthweights")
```

After running the histogram we observe a bell-shaped form, which
indicates that the data is normally distributed. Although we can clearly
see the shape, the histogram alone isn't a good indicator of normally
distributed data.

```{r}
qqnorm(dataset$birthweight, main="Normal Probability Plot of Birthweights")
```

We run the qqnorm function to confirm the distribution. A nearly
straight line indicates that the distribution is indeed normal.

To construct the 96%-CI for the mu birthweight the following
calculations are performed.

```{r}
n <- length(dataset$birthweight)
se <- sd(dataset$birthweight) / sqrt(n)
t_star <- qt(0.98, df=n-1)
ci <- c(mean(dataset$birthweight) - t_star*se, mean(dataset$birthweight) + t_star*se)
```

To get the 96%-CI of the mean we find the sample size (n), standard
error (se), and critical value of t-distribution (t_star). We apply the
formula mū ± t \* (se/√n).

To evaluate the sample size needed to provide that the length of the
96%-CI is at most 100.

```{r}
t_star <- qt(0.98, df=1e6)
n <- ceiling(((t_star * sd(dataset$birthweight) / 50) ^ 2))
```

Here the t_star uses other df value that denotes an arbitrary large
sample size that we need for our calculation.

To compute a bootstrap 96%-CI for mu define a function to generate a
bootstrap sample and compute the mean of the sample.

```{r}
boot_mean <- function(data) {
  boot_sample <- sample(data, replace=TRUE, size=length(data))
  mean(boot_sample)
}
```

Generate 10000 bootstrap samples and store the means in a vector.

```{r}
boot_means <- replicate(10000, boot_mean(dataset$birthweight))
```

Calculate the 2.5th and 97.5th percentiles of the bootstrapped means to
obtain the 96% CI.

```{r}
boot_ci <- quantile(boot_means, c(0.025, 0.975))
```

So the values of the bootstrap CI to the one obtained using the
t-distribution could be compared.

```{r}
ci
boot_ci
```

b)  An expert claims that the mean birthweight is bigger than 2800 gram.
    Verify this claim by using a relevant t-test, explain the meaning of
    the CI in the R-output for this test. Also propose and perform a
    suitable sign tests for this problem.

To verify the claim a one-sample t-test could be performed with null
hypothesis H0: mu = 2800 and alternative hypothesis Ha: mu \> 2800.
Here, mu represents the true population mean birthweight.

```{r}
t.test(dataset, mu=2800, alternative="greater")
```

The output of this test includes the test statistic (t-value), the
p-value, and a confidence interval for the population mean. The p-value
represents the probability of observing a sample mean at least as
extreme as the one we obtained, assuming that the null hypothesis is
true.

The p-value is 0.01357 which is less than the significance level, so we
reject the null hypothesis and conclude that there is strong evidence to
suggest that the mean birthweight is indeed greater than 2800 gram. The
confidence interval also supports this conclusion, as it does not
include 2800 gram.

Alternatively a sign test can be performed to confirm the hypothesis.

```{r}
n <- length(dataset$birthweight)
sign_test <- binom.test(sum(dataset$birthweight > 2800), n, p=0.5, alternative="greater")
sign_test
```

The output of this test will include the test statistic (the number of
observations above 2800 gram), the p-value, and a confidence interval
for the true proportion of observations that are above 2800 gram.
Assuming the output p-value we reject the null hypothesis in favor of
alternative hypothesis.

c)  Propose a way to compute the powers of the t-test and sing test from
    b)  at some mu \> 2800, comment.

To compute the power of the t-test, we need to specify a hypothetical
value of the mean birthweight, denoted by mu1, which is greater than
2800g. Let's say we choose mu1 = 2850g. Then we can use the following
approach to compute the power.

For the t-test, we can compute the t-value and p-value for the null
hypothesis H0: mu = 2800g versus the alternative hypothesis Ha: mu \>
2800g, using the formula: t = (mu -mu0)/(sd/sqrt(n)).

The p-value can then be obtained from the t-distribution with n - 1
degrees of freedom. Once we have the p-value, we can compute the power
of the test for the specified value of mu1.

For the sign test the p-value can be computed for the null hypothesis
H0: P(X \> 2800) = 0.5 against the alternative hypothesis Ha: P(X \>
2800) \> 0.5 using the binomial distribution: p = sum(dbinom(k:n, n,
0.5)),

k - number of observations greater than 2800g.

The p-value is then compared to the significance level α to determine
whether to reject or fail to reject the null hypothesis.

d)  Let p be the probability that birthweight of a newborn baby is less
    than 2600 gram. Using asymptotic normality, the expert computed the
    left end p\^l = 0.25 of the confidence interval [p^l,p^r] for p.
    Recover the whole confidence interval and its confidence level.

If the probability that birthweight of a newborn baby is less than 2600
gram is p, then the number of newborn babies out of 188 with
birthweights less than 2600 gram follows a binomial distribution with
parameters n = 188 and p. Assuming that the sample size is large enough,
we can approximate the distribution of the sample proportion by a normal
distribution with mean p and variance p(1-p)/n.

The expert computed the left end p\^l = 0.25 of the confidence interval
[p\^l, p\^r] for p. Since the normal distribution is symmetric, we have
p\^r = 1 - p\^l = 1 - 0.25 = 0.75.

To recover the whole confidence interval, we need to find the z-value
that corresponds to the left tail probability of 0.25. We can use the
qnorm() function in R to find this value.

```{r}
z_star <- qnorm(0.25)
z_star
```

Using the normal approximation to the binomial distribution, the
confidence interval for p is.

```{r}
p_hat <- sum(dataset$birthweight < 2600) / nrow(dataset)
se_p <- sqrt(p_hat * (1 - p_hat) / nrow(dataset))
ci_p <- p_hat + z_star * se_p + c(0, 1) * 1.96 * se_p
ci_p
```

e)  The expert also reports that there were 34 male and 28 female babies
    among 62 who weighted less than 2600 gram, and 61 male and 65 female
    babies among the remaining 126 babies. The expert claims that the
    mean weight is different for male and female babies. Verify this
    claim by an appropriate test.

To test if the mean birthweight is different for male and female babies,
we can perform a two-sample t-test.

First, we can split the data into two groups based on the gender of the
newborns.
