---
title: "Assignment 1"
author: "Group 69"
date: "24.02.2023"
output: pdf_document
fontsize: 11pt
highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

# Assignment 1

## Exercise 1

## Exercise 2. Cholesterol

*A study tested whether cholesterol was reduced after using a certain
brand of margarine as part of a low fat low cholesterol diet. The data
set
[cholesterol.txt](https://canvas.vu.nl/courses/68030/files/5971196?wrap=1 "cholesterol.txt")contains
information on 18 people using margarine to reduce cholesterol: columns
Before and After8weeks contain the cholesterol level (mmol/L)
respectively before the diet and after 8 weeks on the diet.*

*a)  Make some relevant plots of this data set, comment on normality.
Are there any inconsistencies in the data? Investigate whether the
columns Before and After8weeks are correlated.*

```{r}
data = read.table("cholesterol.txt")
attach(data)
head(data)
```

```{r}
par(mfrow=c(1,2))
qqnorm(Before, main="Before")
qqnorm(After8weeks, main="After")
par(mfrow=c(1,1))
```

Both before and after follow the normal distribution.

```{r}
par(mfrow=c(1,2))
hist(Before, main="Before")
hist(After8weeks, main="After")
par(mfrow=c(1,1))
```

According to the histograms and boxplots, we can see that both Before
and After columns are approximately normal. There are no major
inconsistencies in the data.

```{r}
boxplot(Before, After8weeks, names = c("Before", "After"))
```

```{r}
cor.test(Before, After8weeks)
```

And have a high correlation

```{r}
plot(Before, After8weeks)
```

*b)  Apply two relevant tests (cf. Lectures 2, 3) to verify whether the
diet with low fat margarine has an effect (argue whether the data are
paired or not). Is a permutation test applicable?*

As each row represents the cholesterol level (mmol/L) respectively
before the diet and after 8 weeks on the diet for every person, paired
tests can be conducted.

```{r}
t.test(Before, After8weeks, paired=T)
# symmetric data is required
wilcox.test(Before, After8weeks, paired=T)
```

Both tests showed p values \< .05 which indicates that the we reject
$H_0$ and accept $H_a$ , so **the low fat margarine diet has an effect
on reducing cholesterol levels**.

Additionally, permutation test can be conducted, however due to the
small sample size of 18 entries, test may not be reliable and paired
t-test and Wilcoxon signed-rank test may be more appropriate.

*c)  Let* $X_1,\ldots,X_{18}$ *be the column After8weeks. Assume*
$X_1,\ldots,X_{18}\sim \text{Unif}[3,\theta]$, then use the central
limit theorem to find an estimate $\hat{\theta}$ *for* $\theta$ *and
construct a 95%-CI for* $\theta$*. Can you improve this CI?*

```{r}
data <- After8weeks
x.bar <- mean(data)
theta.hat <- x.bar * 2 - 3
z.val <- qnorm(.975) # z-value for a 95% CI
n <- length(data)
se <- sd(data) / sqrt(n)
CI <- theta.hat - c(1,-1) * z.val * se

theta.hat
CI

```

To improve the CI, we could try using a bootstrap approach to estimate
the distribution of $\hat{\theta}$ and calculate a percentile CI.

Another way is to increase sample size. More realistic way is to change
normal distribution to a t-distribution, as the sample has only 18
records.

*d)  By using a bootstrap test with test statistic*
$T=\max(X_1,\ldots,X_{18})$*, determine those* $\theta\in[3,12]$*, for
which the hypothesis* $X_1,\ldots,X_{18}\sim \text{Unif}[3,\theta]$ *is
not rejected. Can the Kolmogorov-Smirnov test be also applied for this
situation?*

```{r}
n = 18
t = max(After8weeks)
nRep = 100

unifTest = function(thetaMax) {
  tstar = numeric(nRep)
  for (i in 1:nRep) {
    tstar[i] = max(runif(18, 3, thetaMax))
  }
  # manually calculate p-value
  pl=mean(tstar<t)
  pr=mean(tstar>t)
  p=2*min(pl,pr)
  p
}

# generate the sequence of possible thetas with the precision 0.001
# in a given range [3,12]
thetas = seq(3,12,.001)

pvalues = numeric(length(thetas))
for (i in 1:length(thetas)) {
  thetaMax = thetas[i]
  # get p-value of the bootstrap test
  pvalues[i] = unifTest(thetaMax)
}

# retrieve thetas that did not reject H0 of the bootstrap test
# and calculate the range
range(thetas[pvalues > .05])

```

For these $\theta$ ranges the hypothesis
$X_1,\ldots,X_{18}\sim \text{Unif}[3,\theta]$ is not rejected (p value
\> .05).

It is also possible to determine the range using Kolmogorov-Smirnov
test.

```{r}
testIfUnif = function(thetaMax) {
  ks.test(After8weeks, 'punif', 3, thetaMax)$p.value
}

thetas = seq(3,12,.001)
pvalues = numeric(length(thetas))
for (i in 1:length(thetas)) {
  thetaMax = thetas[i]
  pvalues[i] = testIfUnif(thetaMax)
}

range(thetas[pvalues > .05])
```

With a similar test setup we end up getting more precise $\theta$ range.

*e)  Using an appropriate test, verify whether the median cholesterol
level after 8 weeks of low fat diet is less than 6. Next, design and
perform a test to check whether the fraction of the cholesterol levels
after 8 weeks of low fat diet less than 4.5 is at most 25%.*

In order to test median, one can use wilcoxon signed-rank test.

```{r}
wilcox.test(After8weeks, mu = 6, alternative="l", exact = T)
```

Based on the output, we can see that the p-value is 0.223, which is
greater than 0.05. Therefore, we fail to reject the null hypothesis that
that the median cholesterol level is $\ge$ 6. We also do not have enough
evidence to conclude that that the median cholesterol level after 8
weeks of low fat diet is less than 6.

Let us take another approach and test the hypothesis with the binomial
test.

```{r}
n = length(After8weeks)
xSuccess = sum(After8weeks > 6)
binom.test(xSuccess, n, .5, alternative = 'l')
```

$H_0$ is not rejected. The confidence interval of the true proportion is
[0.0000000 0.6078447] which does not help in this case, as it includes
0.5 value.

```{r}
n = length(After8weeks)
xSuccess = sum(After8weeks < 4.5)
binom.test(xSuccess, n, .25, alternative = 'l')
```

We fail to reject the null hypothesis (p\>0.3057) that the true
proportion of cholesterol levels less than 4.5 is 0.25, and do not have
evidence to conclude that the true proportion is less than 0.25.

The 95% confidence interval for the true proportion is [0.0000000
0.3766792], which includes the value 0.25, providing further support for
the null hypothesis. Therefore, we do not have strong evidence to
conclude that the fraction of cholesterol levels after 8 weeks of
low-fat diet less than 4.5 is at most 25%.

## Exercise 3

## Exercise 4
