---
title: "Assignment 1"
author: "Group 69"
date: "24.02.2023"
output: pdf_document
fontsize: 11pt
highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

# Assignment 1

## Exercise 1

## Exercise 2. Cholesterol

*A study tested whether cholesterol was reduced after using a certain
brand of margarine as part of a low fat low cholesterol diet. The data
set
[cholesterol.txt](https://canvas.vu.nl/courses/68030/files/5971196?wrap=1 "cholesterol.txt")contains
information on 18 people using margarine to reduce cholesterol: columns
Before and After8weeks contain the cholesterol level (mmol/L)
respectively before the diet and after 8 weeks on the diet.*

*a)  Make some relevant plots of this data set, comment on normality.
Are there any inconsistencies in the data? Investigate whether the
columns Before and After8weeks are correlated.*

```{r}
data = read.table("cholesterol.txt")
attach(data)
head(data)
```

```{r}
par(mfrow=c(1,2))
qqnorm(Before, main="Before")
qqnorm(After8weeks, main="After")
par(mfrow=c(1,1))
```

Both before and after follow the normal distribution.

```{r}
par(mfrow=c(1,2))
hist(Before, main="Before")
hist(After8weeks, main="After")
par(mfrow=c(1,1))
```

According to the histograms and boxplots, we can see that both Before
and After columns are approximately normal. There are no major
inconsistencies in the data.

```{r}
boxplot(Before, After8weeks, names = c("Before", "After"))
```

```{r}
cor(Before, After8weeks)
```

And have a high correlation

```{r}
plot(Before, After8weeks)
```

*b)  Apply two relevant tests (cf. Lectures 2, 3) to verify whether the
diet with low fat margarine has an effect (argue whether the data are
paired or not). Is a permutation test applicable?*

As each row represents the cholesterol level (mmol/L) respectively
before the diet and after 8 weeks on the diet for every person, paired
tests can be conducted.

```{r}
t.test(Before, After8weeks, paired=T)
wilcox.test(Before, After8weeks, paired=T)
```

Both tests showed p values \< .05 which indicates that the we reject
$H_0$ and accept $H_a$ , so **the low fat margarine diet has an effect
on reduing cholesterol levels**.

Additionally, permutation test can be conducted, however due to the
small sample size of 18 entries, test may not be reliable and paired
t-test and Wilcoxon signed-rank test may be more appropriate.

<!--# Should we make a permutation test? -->

*c)  Let* $X_1,\ldots,X_{18}$ *be the column After8weeks. Assume*
$X_1,\ldots,X_{18}\sim \text{Unif}[3,\theta]$, then use the central
limit theorem to find an estimate $\hat{\theta}$ *for* $\theta$ *and
construct a 95%-CI for* $\theta$*. Can you improve this CI?*

```{r}
x_bar <- mean(After8weeks)
s <- sd(After8weeks)

# calculate estimate of theta and 95% CI
theta_hat <- x_bar + qnorm(0.975) * s / sqrt(length(After8weeks))
ci <- c(x_bar - qnorm(0.975) * s / sqrt(length(After8weeks)), theta_hat)

# print results
cat("Estimate of theta:", theta_hat, "\n")
cat("95% CI for theta:", ci, "\n")
```

To improve the CI, we could try using a bootstrap approach to estimate
the distribution of $\hat{\theta}$ and calculate a percentile CI.

*d)  By using a bootstrap test with test statistic*
$T=\max(X_1,\ldots,X_{18})$*, determine those* $\theta\in[3,12]$*, for
which the hypothesis* $X_1,\ldots,X_{18}\sim \text{Unif}[3,\theta]$ *is
not rejected. Can the Kolmogorov-Smirnov test be also applied for this
situation?*

*e)  Using an appropriate test, verify whether the median cholesterol
level after 8 weeks of low fat diet is less than 6. Next, design and
perform a test to check whether the fraction of the cholesterol levels
after 8 weeks of low fat diet less than 4.5 is at most 25%.*

```{r}
wilcox.test(Before, After8weeks, paired=T, alternative = "less", conf.int=T)
```

A p-value of 1 indicates that the observed data is very likely under the
null hypothesis, meaning that there is no significant difference between
the median cholesterol levels before and after the low-fat diet.

<!--# there is no significat evidence to reject the null hypothesis, which is that cholesterol level after 8 weeks of low fat diet is less than 6. -->

The confidence interval (-Inf, 0.705) suggests that the true difference
in median cholesterol levels between before and after the diet could be
as low as negative infinity (which is not possible) and as high as
0.705. However, since the p-value is not significant, this confidence
interval should be interpreted with caution.

The sample estimate of the (pseudo)median is 0.635, which suggests that
the median cholesterol level after the low-fat diet is slightly lower
than before, but as the p-value is not significant, we cannot conclude
that this difference is statistically significant.

```{r}
less_than_4.5 <- sum(After8weeks < 4.5)
binom.test(less_than_4.5, length(After8weeks), p = 0.25, alternative = "less")
```

The p-value of 0.3057 suggests that there is no significant evidence to
reject the null hypothesis, which in this case is that the fraction of
cholesterol levels after the low-fat diet is less than 4.5 is at most
25%.

The confidence interval of (0.0000000, 0.3766792) suggests that the true
fraction of cholesterol levels after the low-fat diet less than 4.5
could range from 0 (which is not possible) to 0.3766792. Since the
p-value is not significant, this confidence interval should be
interpreted with caution.

The probability of success of 0.1666667 indicates that only 1 out of the
6 participants had a cholesterol level after the low-fat diet less than
4.5, which is a low probability of success. However, as the p-value is
not significant, we cannot conclude that this difference is
statistically significant.

## Exercise 3

## Exercise 4
