---
title: "Assignment 1"
author: "Group 69"
date: "24.02.2023"
output: pdf_document
fontsize: 11pt
highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

# Assignment 1

## Exercise 1

## Exercise 2. Cholesterol

*A study tested whether cholesterol was reduced after using a certain
brand of margarine as part of a low fat low cholesterol diet. The data
set
[cholesterol.txt](https://canvas.vu.nl/courses/68030/files/5971196?wrap=1 "cholesterol.txt")contains
information on 18 people using margarine to reduce cholesterol: columns
Before and After8weeks contain the cholesterol level (mmol/L)
respectively before the diet and after 8 weeks on the diet.*

*a)  Make some relevant plots of this data set, comment on normality.
Are there any inconsistencies in the data? Investigate whether the
columns Before and After8weeks are correlated.*

```{r}
data = read.table("cholesterol.txt")
head(data)
```

```{r}
cor(data$Before, data$After8weeks)
```

```{r}
plot(data$Before, data$After8weeks)
```

*b)  Apply two relevant tests (cf. Lectures 2, 3) to verify whether the
diet with low fat margarine has an effect (argue whether the data are
paired or not). Is a permutation test applicable?*

*c)  Let* $X_1,\ldots,X_{18}$ *be the column After8weeks. Assume*
$X_1,\ldots,X_{18}\sim \text{Unif}[3,\theta]$, then use the central limit theorem to find an estimate $\hat{\theta}$
*for* $\theta$ *and construct a 95%-CI for* $\theta$*. Can you improve
this CI?*

*d)  By using a bootstrap test with test statistic
\$T=\\max(X_1,\\ldots,X\_{18})\$, determine those \$\\theta\\in[3,12]\$,
for which the hypothesis* $X_1,\ldots,X_{18}\sim \text{Unif}[3,\theta]$
*is not rejected. Can the Kolmogorov-Smirnov test be also applied for
this situation?*

*e)  Using an appropriate test, verify whether the median cholesterol
level after 8 weeks of low fat diet is less than 6. Next, design and
perform a test to check whether the fraction of the cholesterol levels
after 8 weeks of low fat diet less than 4.5 is at most 25%.*

## Exercise 3

## Exercise 4
