---
title: "Assignment 1"
author: "Group 69"
date: "24.02.2023"
output: pdf_document
fontsize: 11pt
highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

To investigate the effect of 3 types of diet, 78 persons were divided
randomly in 3 groups, the first group following diet 1, second group
diet 2 and the third group diet 3. Next to some other characteristics,
the weight was measured before diet and after 6 weeks of diet for each
person in the study. The collected data is summarized in the data frame
diet.txt in the following columns: person -- participant number, gender
-- gender (1 = male, 0 = female), age -- age (years), height -- height
(cm), preweight -- weight before the diet (kg), diet -- the type of diet
followed, weight6weeks -- weight after 6 weeks of diet (kg). Compute and
add to the data frame the variable weight.lost expressing the lost
weight, to be used as response variable.

```{r}
diet = read.table("diet.txt", header = T)
diet$weight.lost <- diet$preweight - diet$weight6weeks
```

a)  Make an informative graphical summary of the data relevant for study
    of the effect of diet on the wight loss. By using only the columns
    preweight and weight6weeks, test the claim that the diet affects the
    weight loss. Check the assumptions of the test applied.

To create a graphical summary, we can use a scatter plot of preweight vs
weight6weeks, with the points colored by diet group. To test whether the
diet affects weight loss, we can use a paired t-test on the difference
between preweight and weight6weeks for each participant, with diet group
as a grouping variable.

```{r}
library(ggplot2)

# Scatter plot of preweight vs weight6weeks, colored by diet group
library(ggplot2)

ggplot(diet, aes(x = preweight, y = weight6weeks, color = factor(diet))) +
  geom_point() +
  labs(title = "Weight loss by diet group",
       x = "Pre-diet weight (kg)",
       y = "Weight after 6 weeks (kg)",
       color = "Diet group")

```

To check the assumptions of the paired t-test, we can inspect the
normality of the differences using a histogram and a normal probability
plot:

```{r}
# Histogram of the differences
hist(diet$preweight - diet$weight6weeks, main="Histogram of Weight Loss Differences")

# Normal probability plot of the differences
qqnorm(diet$preweight - diet$weight6weeks, main="Normal Probability Plot of Weight Loss Differences")
qqline(diet$preweight - diet$weight6weeks)
```

The histogram and normal probability plot of the differences show a
roughly symmetric distribution, so the assumption of normality is met.
The variances of the differences are roughly equal across diet groups,
so the assumption of homogeneity of variances is met. If both
assumptions are met, then the paired t-test can be used to test for
significant differences in weight loss between diet groups.

H_0: There is no significant difference in weight loss between the three
types of diets. H_1: There is a significant difference in weight loss
between three types of diets.

```{r}
# Paired t-test
t.test(diet$preweight, diet$weight6weeks, paired=TRUE, alternative="two.sided", var.equal=TRUE)
```

With p-value being smaller than 0.05 we can reject null hypothesis.

b)  Apply one-way ANOVA to test whether type of diet has an effect on
    the lost weight. Do all three types diets lead to weight loss? Which
    diet was the best for losing weight? Can the Kruskal-Wallis test be
    applied for this situation?

To test whether the type of diet has an effect on weight loss, we will
use a one-way ANOVA with diet as a factor. H_0: The means of the
different groups are the same H_1: At least one sample mean is not equal
to the others.

```{r}

#one-way anova
# factor() as TukeyHSD() requires the aov object to have been created with groups as explicit factors
model_anova <- aov(weight.lost ~ factor(diet), data = diet)
summary(model_anova)

#Check the assumptions of the ANOVA
#Residuals vs. Fitted
plot(model_anova, 1)
#Normal Q-Q Plot
plot(model_anova, 2)
#Scale-Location
plot(model_anova, 3)
#Residuals vs. Leverage
plot(model_anova, 5)

```

The ANOVA test shows a significant difference in weight lost between
diet groups as the p-value \< 0.05 and we can reject the null
hypothesis. The assumptions of the ANOVA seem to be met, with the
residuals fairly evenly distributed and no obvious pattern in the
residual plots. F = 6.197. The larger the F value, the more likely it is
that the variance associated with the independent variable is real and
cannot be explained by chance.

An ANOVA tells you whether there are differences between the groups of
the independent variable, but not which differences are significant. To
see how the groups differ from each other, perform a Tukey 's
Honestly-Significant Difference (Tukey HSD) post-hoc analysis.

```{r}
# Tukey's HSD test
TukeyHSD(model_anova)
```

The Tukey test compares the groups pairwise and uses a conservative
error estimate to find the groups that are statistically different from
each other. Results provide the mean difference between each treatment
(diet), the lower and upper bounds of the 95% confidence interval and
the p-value corrected for multiple pairwise equations. As we can see,
the difference between the average result of diet 2 and diet 1 is not
significantly different, while there is a significant difference between
diets 1 and 3 and diets 2 and 3 results.

Kruskal-Wallis test. Technically the Kruskal-Wallis test can be applied
in this situation as an alternative to the one-way ANOVA if the
assumptions of normality and homogeneity of variances are not met.
However, based on the graphical summary of the data and the normal
probability plot of the residuals from the ANOVA model, the assumptions
seem to be reasonably satisfied, so the ANOVA is more appropriate in
this situation.

c)  Use two-way ANOVA to investigate effect of the diet and gender (and
    possible interaction) on the lost weight.

To perform a two-way ANOVA we will use weight.lost as the response
variable, diet and gender as the factors, and their interaction. It is
possible that the effect of diet on weight loss differs between males
and females, or that the effect of gender on weight loss differs
depending on the type of diet. Therefore, this model allows for
different effects of diet and gender on weight loss and also for an
interaction effect between diet and gender, which can be tested for
significance.

H0: There is no significant difference in weight lost between the three
types of diet. H1: There is a significant difference in weight lost
between at least two of the three types of diet.

H0: There is no significant difference in weight lost between males and
females. H1: There is a significant difference in weight lost between
males and females.

H0: There is no significant interaction effect between diet and gender
on weight lost. H1: There is a significant interaction effect between
diet and gender on weight lost.

The p-values obtained from the ANOVA table will allow us to determine
whether to reject or fail to reject each of the null hypotheses. If the
p-value is less than the significance level of 0.05, we reject the null
hypothesis and conclude that there is a significant effect. Factors are
used to represent categorical data.

```{r}
model <- aov(weight.lost ~ factor(diet) + factor(gender) + factor(diet):factor(gender), data = diet)
summary(model)
```

The results of the two-way ANOVA suggest that the type of diet has a
significant effect on the lost weight with the significance level being
0.05 (p-value = 0.00215). At the same time the effect of gender is not
significant (p-value = 0.15987). The interaction between diet and gender
is significant (p-value = 0.04678).

e)  Which of the two approaches, the one from b) or the one from c), do
    you prefer? Why? For the preferred model, predict the lost weight
    for all three types of diet.

The answer depends on the research question and the hypothesis being
tested. If the research question is solely focused on the effect of diet
on weight loss, then a one-way ANOVA for diet would be more appropriate.
However, in this assignment the question was formulated a bit broader,
so I would say that we prefer to pick a two-ways ANOVA as it provides
more insightful result for our case. We consider stating the interaction
factor between gender and diet valuable for the research question and
consequently for hypothesis.

```{r}
model <- aov(weight.lost ~ factor(diet) + factor(gender) + factor(diet):factor(gender), data = diet)

# new data frame for predictions
new_data <- data.frame(diet = factor(c(1, 2, 3)),
                       gender = factor(c(rep(0, 3), rep(1, 3))),
                       diet_gender = factor(c("1_0", "2_0", "3_0", "1_1", "2_1", "3_1"), 
                                            levels = levels(interaction(diet$diet, diet$gender))))

# predict lost weight for each combination of diet and gender
predicted_weight_lost <- predict(model, new_data)

predicted_weight_lost
```

Not mandatory, prediction for the one-way ANOVA just for comparison.

```{r}
model <- aov(weight.lost ~ factor(diet), data = diet)

# Predict the lost weight for all three diets
predictions <- predict(model, newdata = data.frame(diet = c("1", "2", "3")))

# Print the predicted values
predictions
```
